---
title: 'Supplemental Material for _Complementarity in Alliances: How strategic compatibility and hierarchy promote efficient cooperation in international security_'
author:
  - name: J Andres Gannon
format:
  pdf:
    fig-format: png
    number-sections: false
    code-line-numbers: true
    fontsize: "12"
    fig-dpi: 600
    fig-pos: "H"
    keep-tex: false
    include-in-header: 
     text: |
       \usepackage{fvextra}
       \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
       \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
editor: source
execute:
  echo: true
  warning: true
  message: true
  error: true
  cache: true
header-includes:
- \usepackage{tikz}
- \usepackage{pgfplots}
- \pgfplotsset{compat=newest}
- \usetikzlibrary{plotmarks}
- \usetikzlibrary{arrows.meta}
- \usepgfplotslibrary{patchplots}
- \usepackage{grffile}
- \usepackage{caption}
- \usepackage[utf8]{inputenc}
- \usepackage[doublespacing]{setspace}
- \AtBeginEnvironment{tabular}{\singlespacing}
- \usepackage{float}
- \usepackage{multirow}
- \usepackage{tablefootnote}
- \usepackage{pifont}
- \usepackage{newunicodechar}
- \usepackage{booktabs}
- \usepackage{siunitx}
- \newcolumntype{d}{
    S[input-open-uncertainty=,
    input-close-uncertainty=,
    parse-numbers = false,
    table-align-text-pre=false,
    table-align-text-post=false]}
- \usepackage{tabularx}
- \newunicodechar{✓}{\ding{51}}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{wrapfig}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
- \usepackage{graphicx}
- \usepackage{tabularray}
- \UseTblrLibrary{booktabs}
---

```{r knitr-options, cache = FALSE}

required_packages <- c(
  "DoubleML", "bayestestR", "brms", "data.table", "dplyr", "emmeans", "fixest",
  "ggVennDiagram", "ggdist", "ggokabeito", "ggplot2", "ggtext", "here",
  "kableExtra", "knitr", "lgr", "lmerTest", "marginaleffects", "mlr3", 
  "mlr3learners", "modelsummary", "ordbetareg", "rio", "tidybayes", "tidyr", "tinytable"
)

# 'parallel' is part of base R and does not require installation

# Install missing packages
installed <- required_packages %in% rownames(installed.packages())
if (any(!installed)) {
  install.packages(required_packages[!installed])
}

# Load all required packages
lapply(required_packages, library, character.only = TRUE)
options(mc.cores = parallel::detectCores())
```

```{r load data}
df_raw <- readRDS(file = paste0(here::here(), "/data/rDMC_raw_v1.rds"))
df_long <- readRDS(file = paste0(here::here(), "/data/rDMC_long_v1.rds"))
df_wide <- readRDS(file = paste0(here::here(), "/data/rDMC_wide_v1.rds"))

df <- readRDS(file = paste0(here::here(), "/data/03_df-full.rds"))
```

\beginsupplement

\tableofcontents

\newpage

# A. Descriptive statistics

```{r summstats}
#| label: tbl-summstats
#| tbl-cap: "Summary statistics of model variables. Year polynomials omitted from table."
#| tbl-position: H

df |>
  dplyr::select('Year' = year,
                'Division of labor' = dol_nicheolap,
                'Strategic compatibility' = strcomp_cinc_total_scaled,
                'Hierarchy' = hier_laplace_scaled,
                'Peacetime coordination' = milinst,
                'Democracy ratio' = demo_polityprop,
                'Proportion contiguous' = dist_contigprop,
                'Maximum distance (log)' = dist_maxlog,
                'Number of rivals (log)' = rivals_log,
                'Alliance members (log)' = members_log,
                'Alliance age (avg)' = allianceage_avg) |>
  modelsummary::datasummary_skim(Title = "Summary statistics of model variables",
                                 histogram = FALSE,
                                 fmt = 2,
                                 output = "tinytable") |>
  tinytable::theme_tt("resize")
```

\newpage

# B. Manuscript model supplementary info

## Ordered beta results table

```{r}
#| results: hide

df_m1 <- df |>
  dplyr::rename(dv = dol_nicheolap,
                strcomp = strcomp_cinc_total_scaled,
                hier = hier_laplace_scaled)

m1_obr <- ordbetareg::ordbetareg(dv ~ 
                                   strcomp + hier + 
                                   milinst + demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg +
                                   (1 | year_norm),
                                 data = df_m1,
                                 cores = 6,
                                 chains = 6,
                                 iter = 2000,
                                 backend = "cmdstanr",
                                 refresh = 0,
                                 file = "ordbeta",
                                 file_refit = "on_change")
```

```{r}
#| label: tbl-obr
#| tbl-cap: "Log odds coefficient estimates. Time dependencies modeled as year cubic splines given computational constraints for year fixed effects."
#| fig.position: 'H'

coefs <- c("b_strcomp" = "Strategic Compatibility",
           "b_hier" = "Hierarchy",
           "b_milinst" = "Peacetime Coordination",
           "b_demo_polityprop" = "Democracy Ratio",
           "b_dist_contigprop" = "Contiguity Ratio",
           "b_dist_maxlog" = "Maximum Distance (log)",
           "b_rivals_log" = "Number of Rivals (log)",
           "b_members_log" = "Number of Members (log)",
           "b_allianceage_avg" = "Alliance Age (avg)")

modelsummary::modelsummary(m1_obr,
                           statistic = "conf.int",
                           exponentiate = TRUE,
                           coef_map = coefs,
                           coef_omit = "Intercept",
                           metrics = "RMSE",
                           output = "tinytable")

# marginaleffects::plot_predictions(m1_obr, condition = "strcomp")
# marginaleffects::plot_predictions(m1_obr, condition = "hier")
```

\newpage

# C. Alternate explanatory variables

## Strategic Compatibility

```{r}
#| label: fig-strcompcompare
#| fig-cap: "Overlap among strategic compatibility criteria. Counts refer to number of dyad-years."
#| fig.position: H

filedirectoryfig2 <- paste0(here::here(), '/data/dyad_threat.rds')
dyad_threat <- readRDS(file = filedirectoryfig2) |>
  dplyr::mutate(rival_peace = dplyr::if_else(peacelevel < 0.25 | (peacelevel == 0.25 & atop_defense_type == "none"), 1, 0),
                rival_prd = dplyr::if_else(prd_threat == 1 & sscore_below == 1 & atop_defense_type == "none", 1, 0)) |>
  dplyr::mutate(rival_strat = dplyr::if_else(rival_strat == 1, dplyr::row_number(), NA),
                rival_peace = dplyr::if_else(rival_peace == 1, dplyr::row_number(), NA),
                rival_prd = dplyr::if_else(rival_prd == 1, dplyr::row_number(), NA))

ggVennDiagram::ggVennDiagram(list('(1) Strategic rival' = dyad_threat$rival_strat,
                                  '(2) Peace scale rival' = dyad_threat$rival_peace,
                                  '(3) Politically relevant threat environment' = dyad_threat$rival_prd),
                             set_color = c("#1B9E77", "#D95F02", "#7570B3"),
                             label_percent_digit = 1,
                             label_size = 3.5) +
  scale_fill_distiller(palette = "Purples", direction = 1) +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "Dyad-years per strategic compatibility criteria") +
  scale_x_continuous(expand = expansion(mult = .3)) +
  theme(legend.position = "none")
```

```{r}
#| label: tbl-altstrcomp
#| tbl-cap: "Coefficient estimates with alternate variables for strategic compatibility."
#| tbl.position: H

atop_sscore <- rio::import(paste0(here::here(), "/inst/extdata/atop-sscore.csv")) |>
  dplyr::filter(year >= 1970) |>
  dplyr::select(year, ccode1, ccode2, s_un_atop, s_wt_atop, kappa_atop, pi_atop)

atop_cy <- rio::import(paste0(here::here(), "/inst/extdata/atop5_1m.csv")) %>%
  dplyr::filter(yrexit > 1970 |
                  yrexit == 0) %>%
  dplyr::mutate(yrexit = dplyr::if_else(yrexit == 0,
                                        2023,
                                        yrexit)) %>%
  dplyr::select(atopid, member, yrent, yrexit) %>%
  dplyr::distinct() %>%
  dplyr::mutate(row = dplyr::row_number()) %>%
  dplyr::group_by(row, atopid, member) %>%
  dplyr::reframe(year = seq(yrent, yrexit)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-row) %>%
  dplyr::rename(ccode1 = 'member') %>%
  dplyr::filter(year %in% c(1970:2014))

atop_sscoreagg <- dplyr::left_join(atop_cy, atop_sscore) %>%
  dplyr::summarise(strcomp_atopunw = mean(s_un_atop),
                   strcomp_atopwt = mean(s_wt_atop),
                   strcomp_atopkappa = mean(kappa_atop),
                   strcomp_atoppi = mean(pi_atop),
                   .by = c(atopid, year)) %>%
  dplyr::mutate(atopid = as.character(atopid),
                year = as.numeric(year))

df_m1 <- df %>%
  dplyr::mutate(atopid = as.character(atopid)) %>%
  dplyr::left_join(., atop_sscoreagg) %>%
  dplyr::mutate(atopid = as.factor(atopid)) %>%
  dplyr::rename(dv = dol_nicheolap,
                hier = hier_laplace_scaled)

models <- list(fixest::feols(dv ~ 
                               strcomp_atopwt + hier + 
                               demo_polityprop + dist_contigprop + dist_maxlog + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp_atopunw + hier + 
                               demo_polityprop + dist_contigprop + dist_maxlog + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1))

coefs <- c("strcomp_atopwt" = "S-score (weighted)",
           "strcomp_atopunw" = "S-score (unweighted)",
           "hier" = "Hierarchy",
#           "milinst" = "Peacetime Coordination",
           "demo_polityprop" = "Democracy Ratio",
           "dist_contigprop" = "Contiguity Ratio",
           "dist_maxlog" = "Maximum Distance (log)",
#           "rivals_log" = "Number of Rivals (log)",
           "members_log" = "Number of Members (log)",
           "allianceage_avg" = "Alliance Age (avg)")

modelsummary::modelsummary(models, 
                           stars = c('*' = .05, '**' = .01, '***' = 0.001),
                           coef_map = coefs,
                           coef_omit = "Intercept", 
                           gof_map = c("nobs", "r.squared", 'adj.r.squared', 'aic', 'bic', 'rmse'),
                           notes = list('All models include alliance-clustered standard errors.'),
                           output = "tinytable") |>
  tinytable::theme_tt("resize")
```

\newpage

## Hierarchy

```{r}
#| label: tbl-althierarchy
#| tbl-cap: "Coefficient estimates with alternate variables for hierarchy."
#| tbl.position: H

df_m1 <- df |>
  dplyr::rename(dv = dol_nicheolap,
                strcomp = strcomp_cinc_total_scaled,
                hier = hier_laplace_scaled)

models <- list(fixest::feols(dv ~ 
                               strcomp + PotMilCapacity.score + 
                               demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp + Depth.score + 
                               demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp + milinst + 
                               demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp + latent.depth.mean + 
                               demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp + cinc_dispersal + 
                               demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp + cinc_asymmetry + 
                               demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp + gp_ratio + 
                               demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1))

coefs <- c("strcomp" = "Strategic Compatibility",
           "PotMilCapacity.score" = "Potential Military Capacity",
           "Depth.score" = "Alliance Depth",
           "milinst" = "Peacetime Coord",
           "latent.depth.mean" = "Latent Depth",
           "cinc_dispersal" = "CINC Dispersal",
           "cinc_asymmetry" = "CINC Asymmetry",
           "gp_ratio" = "Great power ratio",
           "demo_polityprop" = "Democracy Ratio",
           "dist_contigprop" = "Contiguity Ratio",
           "dist_maxlog" = "Maximum Distance (log)",
           "rivals_log" = "Number of Rivals (log)",
           "members_log" = "Number of Members (log)",
           "allianceage_avg" = "Alliance Age (avg)")

modelsummary::modelsummary(models, 
                           stars = c('*' = .05, '**' = .01, '***' = 0.001),
                           coef_map = coefs,
                           coef_omit = "Intercept", 
                           gof_map = c("nobs", "r.squared", 'adj.r.squared', 'aic', 'bic', 'rmse'),
                           notes = list('All models include alliance-clustered standard errors.'),
                           output = "tinytable") |>
  tinytable::theme_tt("resize")
```

\newpage

## Interaction term

```{r}
#| label: tbl-interaction
#| tbl-cap: "Coefficient estimates with interaction term."
#| tbl.position: H

df_m1 <- df |>
  dplyr::rename(dv = dol_nicheolap,
                strcomp = strcomp_cinc_total_scaled,
                hier = hier_laplace_scaled)

models <- list(fixest::feols(dv ~ strcomp + hier + 
                               strcomp * hier +
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ strcomp + hier + 
                               strcomp * hier +
                               milinst + demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ strcomp + hier +
                               strcomp * hier
                             | year_norm, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ strcomp + hier + 
                               strcomp * hier +
                               milinst + demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg 
                             | year_norm, 
                             cluster = "atopid", 
                             data = df_m1),
               lmerTest::lmer(dv ~ strcomp + hier +
                               strcomp * hier +
                                (1 | year_norm),
                              data = df_m1),
               lmerTest::lmer(dv ~ strcomp + hier + 
                               strcomp * hier +
                                milinst + demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg +
                                (1 | year_norm),
                              data = df_m1))

coefs <- c("strcomp" = "Strategic Compatibility",
           "hier" = "Hierarchy",
           "strcomp:hier" = "Strat Comp*Hierarchy",
           "demo_polityprop" = "Democracy Ratio",
           "dist_contigprop" = "Contiguity Ratio",
           "dist_maxlog" = "Maximum Distance (log)",
           "rivals_log" = "Number of Rivals (log)",
           "members_log" = "Number of Members (log)",
           "allianceage_avg" = "Alliance Age (avg)")

modelsummary::modelsummary(models,
                           stars = c('*' = .05, '**' = .01, '***' = 0.001),
                           coef_map = coefs,
                           coef_omit = "Intercept",
                           gof_map = c("nobs", "r.squared", 'adj.r.squared', 'aic', 'bic', 'rmse'),
                           notes = list('All models include alliance clustered standard errors.'),
                           output = "tinytable") |>
  tinytable::group_tt(j = list("Year Polynomials" = 2:3, "Year FE" = 4:5, "Multilevel Model" = 6:7)) |>
  tinytable::theme_tt("resize")
```

\newpage

# D. Alternate and additional controls

## Alternate controls for democracy

```{r}
#| label: tbl-altdemo
#| tbl-cap: "Coefficient estimates with alternate democracy variables."
#| tbl.position: H

df_m1 <- df |>
  dplyr::rename(dv = dol_nicheolap,
                strcomp = strcomp_cinc_total_scaled,
                hier = hier_laplace_scaled)

models <- list(fixest::feols(dv ~ 
                               strcomp + hier + 
                               demo_polityavg + milinst + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp + hier + 
                               demo_politylowest + milinst + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp + hier + 
                               demo_politydiff + milinst + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp + hier + 
                               demo_sameregime + milinst + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp + hier + 
                               demo_udsavg + milinst + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp + hier + 
                               demo_udslowest + milinst + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp + hier + 
                               demo_udsdiff + milinst + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1))

coefs <- c("strcomp" = "Strategic Compatibility",
           "hier" = "Hierarchy",
           "demo_polityprop" = "Democracy Ratio",
           "demo_polityavg" = "Polity (avg)",
           "demo_politylowest" = "Polity (lowest)",
           "demo_politydiff" = "Polity (difference)",
           "demo_sameregime" = "Same regime type",
           "demo_udsavg" = "UDS (avg)",
           "demo_udslowest" = "UDS (lowest)",
           "demo_udsdiff" = "UDS (difference)",
           "milinst" = "Peacetime Coordination",
           "dist_contigprop" = "Contiguity Ratio",
           "dist_maxlog" = "Maximum Distance (log)",
           "rivals_log" = "Number of Rivals (log)",
           "members_log" = "Number of Members (log)",
           "allianceage_avg" = "Alliance Age (avg)")

modelsummary::modelsummary(models, 
                           stars = c('*' = .05, '**' = .01, '***' = 0.001),
                           coef_map = coefs,
                           coef_omit = "Intercept",
                           gof_map = c("nobs", "r.squared", 'adj.r.squared', 'aic', 'bic', 'rmse'),
                           notes = list('All models include alliance-clustered standard errors.'),
                           output = "tinytable") |>
  tinytable::theme_tt("resize")
```

\newpage

## Alternate controls for geography

```{r}
#| label: tbl-altgeog
#| tbl-cap: "Coefficient estimates with alternate geography variables."
#| tbl.position: H

df_m1 <- df |>
  dplyr::rename(dv = dol_nicheolap,
                strcomp = strcomp_cinc_total_scaled,
                hier = hier_laplace_scaled)

models <- list(fixest::feols(dv ~ 
                               strcomp + hier + 
                               milinst + demo_polityprop + dist_avglog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp + hier + 
                               milinst + demo_polityprop + dist_contigprop + dist_avglog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1))

coefs <- c("strcomp" = "Strategic Compatibility",
           "hier" = "Hierarchy",
           "milinst" = "Peacetime Coordination",
           "demo_polityprop" = "Democracy Ratio",
           "dist_contigprop" = "Contiguity Ratio",
           "dist_maxlog" = "Max Distance (log)",
           "dist_avglog" = "Mean Distance (log)",
           "rivals_log" = "Number of Rivals (log)",
           "members_log" = "Number of Members (log)",
           "allianceage_avg" = "Alliance Age (avg)")

modelsummary::modelsummary(models, 
                           stars = c('*' = .05, '**' = .01, '***' = 0.001),
                           coef_map = coefs,
                           coef_omit = "Intercept",
                           gof_map = c("nobs", "r.squared", 'adj.r.squared', 'aic', 'bic', 'rmse'),
                           notes = list('All models include alliance-clustered standard errors.'),
                           output = "tinytable") |>
  tinytable::theme_tt("resize")
```

## Additional controls for NATO, US, and Cold War

```{r}
#| label: tbl-morectrls
#| tbl-cap: "Coefficient estimates with additional control variables."
#| tbl.position: H

country_year <- rio::import(paste0(here::here(), "/inst/extdata/ATOP/atop5_1m.csv")) %>%
  dplyr::filter(yrexit > 1970 |
                  yrexit == 0) %>%
  dplyr::mutate(yrexit = dplyr::if_else(yrexit == 0,
                                        2023,
                                        yrexit)) %>%
  dplyr::select(atopid, member, yrent, yrexit) %>%
  dplyr::distinct() %>%
  dplyr::mutate(row = dplyr::row_number()) %>%
  dplyr::group_by(row, atopid, member) %>%
  dplyr::reframe(year = seq(yrent, yrexit)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-row) %>%
  dplyr::rename(ccode = 'member') %>%
  dplyr::filter(year %in% c(1970:2014)) |>
  dplyr::filter(ccode == 2) |>
  dplyr::mutate(us = 1) |>
  dplyr::select(atopid, year, us) |>
  dplyr::mutate(atopid = as.factor(atopid)) |>
  dplyr::distinct()

df_m1 <- df %>%
  dplyr::left_join(., country_year) |>
  dplyr::mutate(natowp = dplyr::if_else(atopid %in% c(3180, 3285), 1, 0),
                us = tidyr::replace_na(us, 0),
                coldwar = dplyr::if_else(year <= 1989, 1, 0)) |>
  dplyr::rename(dv = dol_nicheolap,
                strcomp = strcomp_cinc_total_scaled,
                hier = hier_laplace_scaled)

models <- list(fixest::feols(dv ~ 
                               strcomp + hier + 
                               natowp + milinst + demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp + hier + 
                               us + milinst + demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp + hier + 
                               coldwar + milinst + demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + 
                               year_norm + year_sq + year_cube, 
                             cluster = "atopid", 
                             data = df_m1))

# fixest::feols(dv ~ strcomp + hier + demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + year_norm + year_sq + year_cube, cluster = "atopid", split = ~ coldwar, data = df_m1)

coefs <- c("strcomp" = "Strategic Compatibility",
           "hier" = "Hierarchy",
           "milinst" = "Peacetime Coordination",
           "demo_polityprop" = "Democracy Ratio",
           "dist_contigprop" = "Contiguity Ratio",
           "dist_maxlog" = "Maximum Distance (log)",
           "rivals_log" = "Number of Rivals (log)",
           "members_log" = "Number of Members (log)",
           "allianceage_avg" = "Alliance Age (avg)",
           "natowp" = "NATO/Warsaw Pact",
           "us" = "United States member",
           "coldwar" = "Cold War")

modelsummary::modelsummary(models, 
                           stars = c('*' = .05, '**' = .01, '***' = 0.001),
                           coef_map = coefs,
                           coef_omit = "Intercept", 
                           gof_map = c("nobs", "r.squared", 'adj.r.squared', 'aic', 'bic', 'rmse'),
                           notes = list('All models include alliance-clustered standard errors.'),
                           output = "tinytable") |>
  tinytable::theme_tt("resize")
```

\newpage

# E. Other model specifications

## Panel-corrected standard errors

```{r}
#| label: tbl-pcse
#| tbl-cap: "Coefficient estimates using panel-corrected standard errors."
#| tbl.position: H

df_m1 <- df |>
  dplyr::rename(dv = dol_nicheolap,
                strcomp = strcomp_cinc_total_scaled,
                hier = hier_laplace_scaled)

models <- list(fixest::feols(dv ~ 
                               strcomp + hier + 
                               year_norm + year_sq + year_cube, 
                             cluster = c("atopid", "year"), 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp + hier + 
                               milinst + demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg + year_norm + year_sq + year_cube, 
                             cluster = c("atopid", "year"), 
                             data = df_m1),
               fixest::feols(dv ~ 
                               strcomp + hier | 
                               year_norm, 
                             cluster = c("atopid", "year"), 
                             data = df_m1),
               fixest::feols(dv ~ strcomp + hier + 
                               milinst + demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg 
                             | year_norm, 
                             cluster = c("atopid", "year"), 
                             data = df_m1))

coefs <- c("strcomp" = "Strategic Compatibility",
           "hier" = "Hierarchy",
           "milinst" = "Peacetime Coordination",
           "demo_polityprop" = "Democracy Ratio",
           "dist_contigprop" = "Contiguity Ratio",
           "dist_maxlog" = "Maximum Distance (log)",
           "rivals_log" = "Number of Rivals (log)",
           "members_log" = "Number of Members (log)",
           "allianceage_avg" = "Alliance Age (avg)")

modelsummary::modelsummary(models, 
                           stars = c('*' = .05, '**' = .01, '***' = 0.001),
                           coef_map = coefs,
                           coef_omit = "Intercept", 
                           gof_map = c("nobs", "r.squared", 'adj.r.squared', 'aic', 'bic', 'rmse'),
                           notes = list('All models cluster standard errors by alliance and year.'),
                           output = "tinytable") |>
  tinytable::group_tt(j = list("Year Polynomials" = 2:3, "Year Fixed Effects" = 4:5)) |>
  tinytable::theme_tt("resize")
```

\newpage

## Bayesian multi-level model

```{r}
#| results: hide

options(mc.cores = 6, brms.backend = "cmdstanr")
bayes_seed <- 1234

df_m1 <- df |>
  dplyr::rename(dv = dol_nicheolap,
                strcomp = strcomp_cinc_total_scaled,
                hier = hier_laplace_scaled) |>
  dplyr::mutate(atopid = as.factor(atopid))

mlm_full <- brms::brm(brms::bf(dv ~ 
                                 strcomp + hier + 
                                 milinst + demo_polityprop + dist_contigprop + dist_maxlog + rivals_log + members_log + allianceage_avg +
                                 (1 + year_norm | atopid),
                                  decomp = "QR"),
                         data = df_m1,
                         chains = 4, seed = bayes_seed, iter = 4000, threads = brms::threading(2))

# bayestestR::diagnostic_posterior(mlm_full) want an R-hat as close to 1 as possible. If above 1.05 model probably didn't converge
```

::: {#fig-bayesmlm layout="[[ 50, 50 ]]"}
```{r}
#| label: fig-bayesmlm-strcomp
#| fig-cap: "Strategic Compatibility"
#| fig-position: H

mlm_full |> 
  emmeans::emtrends(~ 1,
                    var = "strcomp",
                    at = list(year_norm = 0),
                    epred = TRUE, re_formula = NULL) |>
  tidybayes::gather_emmeans_draws() |>
  ggplot(aes(x = .value)) +
  ggdist::stat_halfeye(fill = ggokabeito::palette_okabe_ito(5),
                       .width = c(0.95)) +
  geom_vline(xintercept = 0) +
  theme_bw() +
  labs(x = "Average marginal effect", 
       y = "Density")
```

```{r}
#| label: fig-bayesmlm-hier
#| results: hide
#| fig-cap: "Hierarchy"
#| fig-position: H

mlm_full |>
  emmeans::emtrends(~ 1,
                    var = "hier",
                    at = list(year_norm = 0),
                    epred = TRUE, re_formula = NULL) |>
  tidybayes::gather_emmeans_draws() |>
  ggplot(aes(x = .value)) +
  ggdist::stat_halfeye(fill = ggokabeito::palette_okabe_ito(5),
                       .width = c(0.95)) +
  geom_vline(xintercept = 0) +
  theme_bw() +
  labs(x = "Average marginal effect", 
       y = "Density")
```

Average marginal effect on division of labor of a one standard deviation increase in the independent variables. Point estimates centered at the median value with 0.95 credible intervals. Note x-axes are not fixed.
:::

\newpage

## Double/debiased Machine-learning

```{r}
#| label: tbl-dml
#| tbl-cap: "Double/debiased machine learning model."
#| fig.position: H

# https://docs.doubleml.org/stable/workflow/workflow.html
library(DoubleML)

# Specify the data and variables for the causal model
df_ml <- data.table::as.data.table(df_m1) |>
  tidyr::drop_na()

dml_data_bonus <- DoubleMLData$new(df_ml,
                                   y_col = "dv",
                                   d_cols = c("strcomp", "hier"),
                                   x_cols = c("demo_udsavg", "dist_maxlog", "dist_contigprop", "rivals_log", "members_log", "allianceage_avg", "year_norm", "year_sq", "year_cube"))

library(mlr3)
library(mlr3learners)
lgr::get_logger("mlr3")$set_threshold("warn")

learner <- lrn("regr.ranger",
               num.trees = 500,
               max.depth = 5,
               min.node.size = 2)

ml_l_bonus <- learner$clone()
ml_m_bonus <- learner$clone()

set.seed(3141)
obj_dml_plr_bonus <- DoubleMLPLR$new(dml_data_bonus,
                                     ml_l = ml_l_bonus,
                                     ml_m = ml_m_bonus)
obj_dml_plr_bonus$fit()

ti <- data.frame(term = c("Strategic compatibility", "Hierarchy"),
                 estimate = obj_dml_plr_bonus$coef,
                 std.error = obj_dml_plr_bonus$se,
                 p.value = obj_dml_plr_bonus$pval)

gl <- data.frame(score = obj_dml_plr_bonus$score,
                 "n-folds" = obj_dml_plr_bonus$n_folds,
                 "cross-fitting" = obj_dml_plr_bonus$apply_cross_fitting)

mod <- list(tidy = ti,
            glance = gl)
class(mod) <- "modelsummary_list"

modelsummary::modelsummary(mod,
                           stars = c('*' = .05, '**' = .01, '***' = 0.001))
```