---
title: '\singlespace A Division of Defense Labor Across Nations'
subtitle: '\singlespace A Theory of the Shared Production of Military Capabilities'
author:
  - name: J Andrés Gannon
    email: "andres.gannon@gmail.com"
    url: jandresgannon.com
    affiliation: Council on Foreign Relations
format:
  pdf:
    fig-format: png
    thanks: The author is thankful to many who provided feedback on earlier and related drafts, including but not limited to Steven Brooks, Rosella Cappella Zielinski, Jonathan Cavereley, Rex Douglass, Ben Fordham, Erik Gartzke, Stacie Goddard, Nadiya Kostyuk, Kendrick Kuo, David Lake, Brett Ashley Leeds, Erik Lin-Greenberg, Paul MacDonald, Michaela Mattes, Steven Miller, Sara Plana, Paul Poast, Abigail Post, Philip Roeder, Sebastian Rosato, Erik Sand, Thomas Scherer, Kaija Schilde, Todd Sechser, Branislav Slantchev, Jennifer Spindel, Sanne Verschuren, and many others. This project could not have happened without help from the many research assistants helped construct the rDMC dataset on which this article relies. This project benefited from financial support from the UC San Diego Center for Peace and Security Studies (cPASS), Harvard Kennedy School Belfer Center for Science and International Affairs, the Smith Richardson Foundation, Stanton Nuclear Foundation, the Department of Defense Minerva Initiative, the Charles Koch Foundation, Eurasia Group Foundation, and the APSA Centennial Center.
    number-sections: true
    fontsize: "12"
editor: visual
execute:
  echo: false
  warning: false
  cache: true
date: today
date-format: long
abstract: \singlespacing \noindent Theories of alliance burden-sharing have primarily investigated the size of allied state militaries, but not their composition. Why do some alliance partners engage in a division of labor over the security capabilities they produce, while other alliance relationships maintain redundant militaries with overlapping capabilities? I argue that alliance relationships can promote an efficient division of labor over the production of defense assets when that alliance relationship has high strategic compatibility and is hierarchical. These two conditions make it easier for states to minimize the risk of defection and ensure effective coordination in a manner that allows them to distribute defense capabilities efficiently across actors. In doing so, states in military alliances can coordinate their defense in a way that garners the benefits of individual specialization and collective diversification. I substantive this argument using data on disaggregated national military capabilities from 1970 -- 2014.
header-includes:
- \usepackage{tikz}
- \usepackage{pgfplots}
- \pgfplotsset{compat=newest}
- \usetikzlibrary{plotmarks}
- \usetikzlibrary{arrows.meta}
- \usepgfplotslibrary{patchplots}
- \usepackage{grffile}
- \usepackage{caption}
- \usepackage[utf8]{inputenc}
- \usepackage[doublespacing]{setspace}
- \AtBeginEnvironment{tabular}{\singlespacing}
- \usepackage{float}
- \usepackage{multirow}
- \usepackage{tablefootnote}
- \usepackage{pifont}
- \usepackage{newunicodechar}
- \usepackage{booktabs}
- \usepackage{siunitx}
- \usepackage{tabularx}
- \newunicodechar{✓}{\ding{51}}
bibliography: rDMC-bib.bib
---

```{r knitr_options}
library(knitr)
library(kableExtra)
library(magrittr)
library(ggplot2)
library(ggtext)
```

```{r load data}
df_raw <- readRDS(file = paste0(here::here(), "/data/rDMC_raw_v1.rds"))
df_long <- readRDS(file = paste0(here::here(), "/data/rDMC_long_v1.rds"))
df_wide <- readRDS(file = paste0(here::here(), "/data/rDMC_wide_v1.rds"))

df <- readRDS(file = paste0(here::here(), "/data/03_df-full.rds"))
```

\newpage

# Introduction

In 1994, Albania and Georgia were among the 18 Former Soviet Union (FSU) and Warsaw Pact states to join the Partnership for Peace, a new program designed to initiate military cooperation with prospective NATO applicants [@szayna_natoenlargement20002015_2001]. Albania and Georgia were similar in many respects like GDP, military spending, geographic size, and coastline. Yet Figure \ref{fig-albania_georgia} shows that by 2014, their militaries looked quite different. Albania specialized in naval patrol vessels with a range that allowed them to patrol as far as Portugal, despite their own coastline being only 225 miles [@polak_natomembershipalbania_2009]. They also omitted air capabilities almost entirely, disposing of all former Soviet combat helicopters and fixed wing aircraft in favor of minimal air power capable only of surveillance and humanitarian missions. Georgia, by comparison, maintained a more diversified military that was similar in size, but more evenly distributed in composition.

```{r}
#| label: fig-albania_georgia
#| fig-cap: "Albania and Georgia specialization."

df_long %>%
  tidyr::separate(tek,
                  into = c("parent", "child"),
                  sep = "_") %>%
  dplyr::group_by(year, country, parent) %>%
  dplyr::summarise(parent_count = sum(unit_count,
                                      na.rm = TRUE),
                   .groups = "keep") %>%
  dplyr::ungroup() %>%
  dplyr::group_by(year, parent) %>%
  dplyr::mutate(parent_ratio = parent_count/sum(parent_count,
                                                na.rm = TRUE)) %>%
  dplyr::ungroup() %>%
  dplyr::filter(year == 2014) %>%
  dplyr::filter(country %in% c("albania", "georgia")) %>%
  dplyr::mutate(country = stringr::str_to_title(country)) %>%
  dplyr::mutate(parent = dplyr::recode(parent,
                                       'principal surface combatants' = 'surface ships',
                                       'patrol and coastal combatants' = 'coastal ships',
                                       'anti-tank/anti-infrastructure' = 'anti-tank',
                                       'armoured fighting vehicles' = 'combat vehicle',
                                       'unmanned aerial vehicles' = 'uav')) %>%
  dplyr::mutate(color = "red") %>%
  dplyr::select(country, parent, parent_ratio) %>%
  dplyr::mutate(parent = stringr::str_to_title(parent)) %>%
  dplyr::mutate(parent = dplyr::if_else(parent == "Uav", "UAV", parent)) %>%
  dplyr::mutate(parent = forcats::fct_relevel(parent, c("Air Defence", "Aircraft", "Helicopters", "UAV", "Ballistic Missiles", "Radars", "Submarines", "Surface Ships", "Coastal Ships", "Amphibious", "Logistics And Support", "Mine Warfare", "Land/Sea Defence", "Anti-Tank", "Combat Vehicle"))) %>%
  dplyr::group_by(country) %>%
  tidyr::spread(key = parent, value = parent_ratio) %>%
  ggradar::ggradar(values.radar = c("", "0.3%", "0.5%"),
                   grid.min = 0,
                   grid.max = 0.005,
                   grid.mid = 0.003,
                   group.line.width = 0.75, 
                   group.point.size = 1.5,
                   plot.extent.x.sf = 1.75,
                   plot.extent.y.sf = 1.25,
                   background.circle.colour = "white",
                   gridline.mid.colour = "grey",
                   gridline.label.offset = 0.001,
                   axis.label.size = 3.5,
                   grid.label.size = 3.5,
                   legend.text.size = 12,
                   legend.position = "top",
                   group.colour = c("forestgreen","orange"))
```

It has long been understood that states are not all like-units.[^1] However, barring information about how states differ, it is tempting to treat them as equivalent, or at least arrayed along a single measurable metric. Doing so has implied that if states had similar levels of resources, they would produce similar amounts of military power.[^2] The result of this assumption is that variation in how states arm themselves is primarily measured in size, rather than composition. However, there are countless examples of different military capabilities which raises the question: why do states possess the force structure that they do [@kurth_whywebuy_1973]? Here, I relax that assumption using new data on the make-up of military capabilities across states from 1970--2014. Doing so shows how states are differentiated not just in the size of their militaries, but in the specific military capabilities in which states have invested.

[^1]: For example, see @onuf_worldourmaking_1989, @watson_evolutioninternationalsociety_1992, @buzan_logicanarchyneorealism_1993, @ruggie_multilateralismmatterstheory_1993, @spruyt_sovereignstateits_1994, @deudney_bindingsovereignsauthorities_1996, @wendt_hierarchyanarchyinformal_1995, @paul_sovereigntysurvivalwestphalian_1999, @lake_newsovereigntyinternational_2003, @lake_escapestatenature_2007, @sharman_internationalhierarchiescontemporary_2013.

[^2]: One theoretical explanation assumption is that socialization and competition under anarchy results in convergence [@waltz_theoryinternationalpolitics_1979; @posen_sourcemilitarydoctrine_1984; @resende-santos_anarchyemulationmilitary_1996; @goldman_systemiceffectsmilitary_1999; @resende-santos_neorealismstatesmodern_2007; @parent_balancingneorealism_2015].

To do this, I focus on the role that alliances play as one part of the explanation for states' strategic decision to choose different distributions of military capabilities even when the structure of the international system and economic capacity are held constant. Why do some pairs of allies specialize their respective military portfolios in a way that complements the capabilities of their ally, while other pairs of allies maintain more redundant militaries?

The institutional form of interstate cooperation helps explain functional differentiation as a conscious policy choice by actors that rely on each other for security. Allied states can engage in division of labor over the production of defense assets -- each specializing in complementary capabilities -- when their relationship is characterized by high strategic compatibility and hierarchy that they expect to continue into the future. This explains variation in the distribution of military capabilities across states not simply as a product of economic capacity and geography, but also as a function of states' willingness and ability to omit and/or overproduce some defense capabilities based on the capabilities of their partners. This builds upon existing theories about the conditions under which major states try to gain security through joint efforts [@jervis_balanceconcertstudy_1986, 58]. I examine the consequences of those joints efforts and add a nuanced understanding to what it means to be 'joint'.

This paper is organized into five sections. @sec-lit outlines existing thinking concerning the degree to which states exist in an anarchic world of 'self-help' concerning defense. @sec-theory then details a theory of a shared production model of defense outlining the conditions under which cooperation under anarchy enables functional differentiation in the production of security -- strategic compatibility and hierarchy. In doing so, it develops a typology that explains the various forms of security cooperation and state military capabilities that should exist based on these two conditions. @sec-empirics provides an empirical test of that theory by identifying the division of labor in security capabilities across all states from 1970 -- 2014 using a novel dataset of disaggregated military capabilities. @sec-conclusion concludes with the implications of these findings for theories about international cooperation and conflict and avenues for future research.

# Existing Explanations for Cooperative Security {#sec-lit}

The previous chapter details existing explanations for the composition of a state's military, ranging from economic and geographic considerations [@brooks_producingsecuritymultinational_2005] to domestic politics [@allison_armamentsarmscontrol_1975; @kehr_battleshipbuildingparty_1975] to social considerations and status [@spinardi_whynavywent_1990; @eyre_statusnormsproliferation_1996]. The prevailing political explanation concerns the nature of the international threat environment [@mcnamara_remarkssecretarydefense_1967; @rathjens_dynamicsarmsrace_1969; @berman_sovietstrategicforces_1982].

If the international system is anarchic and governed only by the logic of self-help, then states can deal with that threat environment through either internal or external balancing [@waltz_theoryinternationalpolitics_1979; @mearsheimer_tragedygreatpower_2001]. The logic of internal balancing means states provide for their own security by arming themselves while external balancing means cooperating with allies against the overarching threat. A result of internal balancing is that state militaries start to look similar overtime as states mimic the capabilities of greater powers, with differences explained by the distribution of resources or geography [@parent_balancingneorealism_2015]. External balancing, when it happens at all, does not change the composition of military capabilities each states possesses because of concerns that cooperation cannot be guaranteed. And in the event that State A and B do cooperate, the stronger of the two will be able to impose its will and preferences over the other rather than engaging in a "division of labor across nations" [@waltz_theoryinternationalpolitics_1979, 105] since self-interested states have no incentive to provide for the security of another state absent that provision enhancing its own security.

Opposition to this neorealist view of anarchy has come from the *hierarchic view of international politics*. If State A and B faced a similar threat environment because of State C, State A and B should engage in a division of labor over the production of security assets to most efficiently provide for their collective security [@lake_entanglingrelationsamerican_1999]. They have everything to gain and very little to lose from that cooperation [@axelrod_achievingcooperationanarchy_1985]. The problem with the previous perspective is thinking about power as a component of competition that all states are engaged in without consideration of the cost of fighting. In reality there are bystanders who gain or lose based on the outcome of competition between two states and the cost of competition can shape whether the gains of competition are worthwhile.[^3] While quoted above in recognizing the best defense against an unpredictable enemy is diversified combined arms, @till_maritimestrategytwenty_1994 [185] also noted that it may be "necessary to augment these by improving access to the capacities of the other services, and of friendly navies."

[^3]: See the discussion on relative and absolute gains by @grieco_anarchylimitscooperation_1988; @snidal_relativegainspattern_1991; @powell_absoluterelativegains_1991.

So why do some capable countries have gaps in their militaries that they could fill, but choose not to, while other capable countries appear dissatisfied with those gaps? This functional differentiation is not just the product of the distribution of power and it does not appear that states are trying to be self-sufficient in their defense [@mawdsley_armamentsdecisionmakingare_2018, 260], as some scholars have predicted [@bitzinger_asianarmsindustries_2017].

And yet, while proponents of this theory would predict a division of labor in defense among like-minded states, the empirical record is far from set. To take Europe as an example, while "one might expect a continent with both a long-established military alliance and highly integrated economies and policymaking machineries to also have a highly integrated defense economy" [@mawdsley_armamentsdecisionmakingare_2018], others have found that Europe's defense market is actually quite fragmented and protectionist [@bitzinger_europeandefenseindustry_2009]. The 27 states in the EU have a combined 25 armies, 21 air forces, and 18 navies most of which possess different weapons systems and that rarely coordinate force planning [@howorth_securitydefencepolicy_2007]. This is not a new problem either; not long after the end of the Cold War [@devestel_defencemarketsindustries_1995] noted that the redundancy of Europe's defense platforms was becoming increasingly costly.[^4] This duplication was not inevitable and could have been ameliorated through pooling and sharing agreements.[^5]

[^4]: @devestel_defencemarketsindustries_1995 measure duplication as the number of platforms and systems in production, rather than the quantities of these platforms, and they are interested in domestic production rather than state ownership and acquisition.

[^5]: Some have argued this is easier to achieve when there is cultural similarity, trust and solidarity, militaries of similar size and strength, and low levels of corruption [@valasek_survivingausteritycase_2011; @briani_armamentsduplicationeurope_2013]. @mawdsley_armamentsdecisionmakingare_2018 rejects this hypothesis because of skepticism that these conditions exist in the European context. While these scholars have looked at arms collaboration, I am here interested in arms distribution -- both within a states and among a collection of states.

So the question becomes why we see a division of labor in some cases but not others and, more interestingly, the form that this division of labor takes. If states facing a common threat feel they are in the world of anarchy, they will have diversified military capabilities since mistrust results in "self-help" defense. But if states facing a common threat have some reason to believe they can effectively cooperate despite anarchy, they will have specialized military capabilities that complement one another since mechanisms for fostering that cooperation enable them to "provide for the common defense". Because my theory borrows generously from this perspective, a comprehensive account of existing research is embedded into the following theory.

# Theory: Bringing balance to each other's force {#sec-theory}

What are the conditions under which two allies are able to ensure that a promise to cooperate with each other in the development and deployment of military capabilities happens successfully? Generally speaking, alliances are a promise to cooperate with another actor under a given set of contingencies [@papayoanou_intraalliancebargainingbosnia_1997]. In the context of military capabilities, a specialized division of labor with an ally can thus be an efficient way to undertake defense cooperation, but it requires bargaining over the terms of that cooperation. Research on bargaining within alliances has started from the @ricardo_principlespoliticaleconomy_1817 model of comparative advantage in trading goods based on differences in production costs [@snyder_conflictnationsbargaining_1977; @snyder_securitydilemmaalliance_1984; @morrow_armsalliestradeoffs_1993; @snyder_alliancepolitics_1997]. One of the things states in an alliance bargain over is which produces what capabilities because these capabilities differ in their asset specificity, economies of scale (private benefits), and contribution to aggregated defense (public benefits). Such bargaining must address the risk of cooperating in the international arena -- the risk of opportunism and costs of coordination.

This theory presents a bargaining model that differentiates it from other conceptions of principal-agent problems in the context of military contracting and arms sales [@devore_armscollaborationdilemma_2011; @borghard_friendsbenefitspower_2014; @berman_proxywarssuppressing_2019; @powell_whypersistentproblems_2019]. When two aligned states bargain over the desired division of labor for the production of security assets, they are both simultaneously serving the roles of principal and agent. I define aligned states as states with whom one could have cooperative security relations where there are expectations of support and/or mutual coordination during future interactions.[^6] Each would like the other to serve as an agent to which they can outsource some aspect of defense.[^7] Specialization is thus a way for states to ensure the arrangement is mutually desirable for both parties by answering the question 'what can you bring to the bargaining table?' The degree to which the arrangement is mutual and, similarly, whether the outsourcing is truly bidirectional is subject to much variation that will be explained in more detail in @sec-hierarchy on hierarchy.

[^6]: Formal military alliances represent one type of alignment, but not all alignments are alliances [@wilkins_alignmentnotalliance_2012].

[^7]: Research on principal-agent relations shares a similar logic concerning delegation. Principals grant power to an agent to gain the benefits of a division of labor. The issue is how the relationship can be controlled and monitored to ensure compliance [@cowhey_delegationnetworksinternet_2009, 173].

Forms of governance differ in how well they enable states to contract to produce pooled security. Yet theories borrowed from industrial organization that originally conceived of governance decisions as being either markets or hierarchies [@williamson_marketshierarchiesanalysis_1975; @perrow_complexorganizationscritical_1986] has since realize there are a variety of governance structures that can produce cooperation to different degrees and in different forms [@williamson_economicinstitutionscapitalism_1985; @powell_neithermarkethierarchy_1990; @heide_interorganizationalgovernancemarketing_1994; @robicheaux_structuremarketingchannel_1994; @zaheer_relationalgovernanceinterorganizational_1995]. In parallel, there has been a wide body of research on the variation of the design of cooperative security structures in interstate relations [@lake_anarchyhierarchyvariety_1996; @weber_hierarchyamidstanarchy_1997; @haftendorn_imperfectunionssecurity_1999; @lake_entanglingrelationsamerican_1999; @weber_hierarchyamidstanarchy_2000; @lake_anarchyimportancesecurity_2001; @leeds_alliancetreatyobligations_2002].

While not theorizing about the origins of these different forms of governance, I here identify two mechanisms by which forms of governance make a division of labor more desirable by simplifying intra-alliance bargaining. The political economy decision about "make or buy" [@coase_naturefirm_1937] serves as a useful analogy to what these forms of governance have in common. Concerning defense, states have a decision about "make/buy or rely". When states "make/buy", they operate in a self-help world of anarchy where they are responsible for their own security. The alternate is engaging in some type of security relationship that allows you to "rely" so that you can forgo the costs of building or buying it yourself [@morrow_armsalliestradeoffs_1993; @digiuseppe_armsdemocraticallies_2016].

This is easier to achieve when the actors' interests are more closely aligned [@papayoanou_intraalliancebargainingbosnia_1997] and when there is hierarchy, allowing one actor to dictate the terms of the bargain [@krasner_globalcommunicationsnational_1991]. If intra-alliance bargaining does not succeed because a gap in strategic compatibility has narrowed the bargaining range and/or because one actor cannot dictate the terms of the bargain sufficient to reach a mutually agreeable solution, then there will not be a shared division of labor over the production of security assets and states will thus design defense portfolios that operate independent of one another; a return to the anarchy-driven combined arms model that seeks individual diversification rather than individual specialization under a structure of collective diversification.

The degree of of strategic compatibility and hierarchy required to overcome this problem is not constant. Two states with similar resource endowments, technological capacity, and geography may both want to specialize in the same set of capabilities given the cost required to do so. For example, two neighboring island states may both want to undertake the naval patrol portion of specialization rather than air defense if their industrial shipbuilding capacity makes that comparatively cheaper. In this case, the degree of division of labor may be lessened by the similarity of both countries.

The fact that both specialization and diversification have distinct benefits poses a dilemma that can only be resolved by intra-alliance bargaining. Even if collective diversification is Pareto optimal, intra-alliance bargaining is necessary to address the temptation to free ride by specializing to such an extent that the other alliance partner is forced to diversify. But since specializing creates reliance on the other actor, mutual specialization now presents a way to resolve the collective action problem by creating a mutual hostage-taking situation [@williamson_crediblecommitmentsusing_1983]. By forcing mutual reliance where both actors' militaries are reasonably hamstrung absent the contribution of the other, each can be assured that the other will not defect because that implicates the shadow of cooperation in the future.[^8]

[^8]: For a division of labor to truly resolve the credible commitment problem, mutual hostage-taking must involve specific assets. If one actor can quickly and easily produce the military capabilities in which the ally specializes, there is little actual reliance. Future research could quantify asset specificity of military technologies by weighing factors like production lead time, although generalizing that across time and space may prove challenging.

These two factors make it easier to reach an intra-alliance bargain by reducing the risk of opportunism and reducing coordination costs. When this happens, states are able to specialize their defense portfolios in ways that garner the efficiency gains of specialized production while maintaining the security gains of a diversified defense portfolio. The result is functionally differentiated military force structures across states in the international system.

## Strategic Compatibility {#sec-interest-alignment}

Strategic compatibility describes the consistency of states' security interests and agreement on the nature of the international threat environment. Whether states have common security interests is a function of whether they face the same threats and the mutual desirability of ways to deal with that security threat. When security interests between two states are consistent, an adversary that poses a threat to one state's security interests also poses a threat to the other state's security interests [@yarhi-milo_armallypatron_2016]. In this situation, states are more likely to have compatible payoff structures regarding actions that should be taken to get the optimal international environment [@axelrod_achievingcooperationanarchy_1985].[^9]

[^9]: A related but distinct aspect of interest alignment is interest intensity. Two actors may have the same interests in that they have a shared understanding of a common goal, but one may be much less willing to expend resources for that goal to be accomplished. In cases where interest intensity diverges significantly, the alignment of interests may be insufficient to encourage cooperation [@gulati_twofacetscollaboration_2012].

If we think about strategic compatibility as a shared understanding of agreed upon goals, then strategic compatibility can be seen as a necessary condition for cooperation [@axelrod_achievingcooperationanarchy_1985; @oye_explainingcooperationanarchy_1985]. However, circumstances where interests are perfectly aligned seem exceedingly rare -- if not non-existent -- in international relations. And unless there is perfect harmony in strategic compatibility, the risk of opportunism can only be reduced, not eliminated, as long as each actor retains autonomy over its own decision-making [@gulati_twofacetscollaboration_2012]. While formal alignments themselves may not often change rapidly or substantially, the salience of the threats that a particular alignment can reliably help counter does change.

For strategic compatibility to incentivize security cooperation, and thus encourage a division of defense labor, it must increase the gains of cooperation and reduce its costs (opportunism and coordination costs). Strategic compatibility improves the gains of cooperation by augmenting the effectiveness of coalition contributions to war [@stueck_koreanwarinternational_1997; @kreps_coalitionsconvenienceunited_2011; @mcinnis_varietiesdefectionstrategies_2019; @cappellazielinski_organizingperformancecoalition_2020]. Closely aligned preferences make crisis bargaining within the coalition easier which improves that coalition's ability to credibly signal resolve. It also positively shapes that coalition's collective strategy in the event that conflict does break out [@wolford_politicsmilitarycoalitions_2015].[^10] Cases of collective security are representative of cases of high strategic compatibility. When a state opts for collective security, they are essentially conveying that the security of the whole is an vital component of the security of the homeland [@conybeare_tripleententetriple_1990].

[^10]: Extensive work has been done on the consequences of preference misalignment in the context of extended deterrence and alliance formation but this has not been applied to the more general context of security cooperation that manifests in the sharing of access to an ally's military assets [@morrow_alliancescredibilitypeacetime_1994; @smith_allianceformationwar_1995; @benson_constructinginternationalsecurity_2012].

High strategic compatibility makes it easier to overcome opportunism and coordination costs because the presence of a common objective that both actors seek produces higher payoffs to conscious policy coordination [@oye_explainingcooperationanarchy_1985; @thies_friendlyrivalsbargaining_2003; @wolford_showingrestraintsignaling_2014]. This is especially true in cases where the common threat facing two states is something like a territorially acquisitive great power since they both have an interest in mutually producing the capacity to respond to that threat. Military specialization has a symbiotic relationship with this end.

For example, the United States and West Germany had aligned security interests during the Cold War because they both saw the Soviet Union as their primary adversary. As a result, the US viewed efforts to enhance West Germany's security as being consistent with the enhancement of its own security [@lanoszka_consentcoercionusing_2013]. In this way, high strategic compatibility encourages cooperation by lowering the risk of opportunism. In these cases, there is a reduced -- albeit, not eliminated -- need for external enforcement mechanisms because the punishment a state faces from defection -- an undesirable threat environment -- is synonymous with failure by your ally. The value of cooperation is what makes it self-enforcing [@keohane_reciprocityinternationalrelations_1986]. As a result, closely aligned interests mean states will contribute to the security of their ally because they have a security incentive to do so even in the event that their partner may defect.

By increasing the expected gains of cooperation and overcoming the expected costs of opportunism and coordination, high strategic compatibility can increase states' willingness to embrace a shared production model of military capabilities. By complementing each other's forces, the gains from economies of scale mean that each state is better off than if they simply added their redundant military capabilities together. The gains from specialization can now be realized if the accompanying costs have been sufficiently reduced by closely aligned interests. Importantly, the rewards of shared production can be reaped internally. Since reducing the production of particular capabilities is one manifestation of specialization, states benefit economically from sharing the burden.

The US relationship with Australia took a significant turn in the early 1970s when Nixon's Guam Doctrine announced US withdrawal from Asia and an expectation that its allies in the region do more to defend themselves [@curran_dilemmasdivergencecrisis_2014]. Australia's concern about the US retreat from Asia after the Vietnam War was pronounced and marked the beginning of a new Australian perspective that they were alone in their defense since their interests were no longer aligned with that of the United States. The Defence White Paper [@commonwealthofaustralia_australiandefence_1976] noted "it is not our policy, nor would it be prudent, to rely upon US combat help in all circumstances". The new self-reliance approach to defense directly influenced force structure decisions since Australia knew any capability that was needed for defense or to achieve international objectives would have to be domestically owned [@fruhling_australiandefencepolicy_2014]. As a result, their military is less specialized than that of others in the region.

*Hypothesis: Defense alliances with high strategic compatibility should have a higher division of labor than alliances without high strategic compatibility.*

Of course, this relationship is endogeneous -- a state's relationship with other states influences the capabilities each state produces but the capabilities each state has at their disposal also impacts the decision to ally with another state [@digiuseppe_armsdemocraticallies_2016]. States may specialize because the omission or surplus production of particular capabilities creates the conditions for mutual vulnerability and interdependence as a form of hostage-taking [@williamson_crediblecommitmentsusing_1983]. But rather than think of this endogeneity as a barrier to casual inference, it instead also explains why this relationship may be enduring. If a state has a demand for a particular military capability that is part of another state's military portfolio, and the first are unable to make or buy that capability, it may strengthen its relationship with the other state so as to enhance the first's ability to borrow [@conybeare_portfoliodiversificationmodel_1992].[^11] If this is true, then the nature of the alignment relationship is still influencing the types of capabilities states are producing. A small and vulnerable state would strategically maintain dependence on a powerful ally by specializing its own military in a way that ensures the powerful ally maintains that relationship.

[^11]: @digiuseppe_armsdemocraticallies_2016 deal with the coterminous relationship between arming and alliances by separately modeling defense pact ties and military expenditure as outcome variables and then jointly estimating them with a conditional mixed-process (CMP) estimator.

## Hierarchy {#sec-hierarchy}

Hierarchy describes the extent to which one of the allies makes decisions about the nature of cooperation in the event of contingencies that had not been anticipated [@zaheer_relationalgovernanceinterorganizational_1995; @donnelly_sovereigninequalitieshierarchy_2006]. I choose the term hierarchy because this variable is about the condition of control and the location of decision-making in intra-alliance bargaining rather than the source of that control or the manner in which it is exercised [@lake_risefallfuture_1997]. While anarchy means that all arrangements about pooling defense efforts are largely self-enforcing, the more hierarchical the arrangement, the more decisions about self-enforcement are controlled by the dominant actor [@jung_marketshierarchiesnetworks_2011].

Hierarchy is not just about the distribution of material power. That concept is too durable and does not give appropriate agency to the actors whose agency creates the relationship [@onuf_worldourmaking_1989; @onuf_makingsensemaking_2013]. While some hierarchical relationships have an unequal distribution of material military power such that "a more powerful state has the material capability to intervene in and provide security for the weaker one" the presence of such a capability is not synonymous with hierarchy nor is its absence indicative of a horizontal relationship [@wendt_hierarchyanarchyinformal_1995, p 696].

Hierarchy increases the relative gains of cooperation by reducing the risk of opportunism and the cost of coordination. It reduces the risk of opportunism through three mechanisms; solving information asymmetries, increasing reputation costs, and creating mutual interdependence. The first of these, information asymmetries, are resolved by hierarchy by providing rules of thumb concerning the role each state plays in the relationship [@oye_explainingcooperationanarchy_1985]. The dominant state delegates nodes of responsibility to the subordinate state either because those tasks are less important niche capabilities or because the subordinate state can perform those tasks at a lower cost given comparative advantage offered by geography or industrial capacity [@sugiyama_improvementrobustnessenvironmental_2017]. This can reduce uncertainty about its costs because you have some idea of how they will act in turn. When that happens in both directions, there is confidence they won't act opportunistically [@axelrod_evolutioncooperation_1984]. By transferring a purely exchange relationship into a power relationship, hierarchy ensures unified command [@galbraith_organizationdesign_1977; @gulati_architecturecooperationmanaging_1998]. NATO did this by explicitly linking the stationing of US troops abroad in exchange for countries purchasing US military equipment [@axelrod_achievingcooperationanarchy_1985].

Reputational costs matter for cooperation because actors are almost always in mutual overlapping alliances or have an expectation of possible alliances in the future [@gulati_mutuallyassuredalliances_1992, p 19]. If international cooperation is a game of repeated play, then actors have to demonstrate that they are worthwhile partners [@tomz_reputationinternationalcooperation_2007]. Hierarchy increases reputational costs by centralizing decisions about issue linkage and creating precision in how cooperation will happen [@abbott_hardsoftlaw_2000; @koremenos_rationaldesigninternational_2001; @mattes_reputationsymmetryalliance_2012]. This reduces the risk of opportunism by creating exit costs to reneging on cooperation [@weber_hierarchyamidstanarchy_1997] which also facilitates reciprocity and further cooperation [@gulati_doesfamiliaritybreed_1995; @malhotra_effectscontractsinterpersonal_2002; @mellewigt_trustformalcontracts_2007]. By making reciprocity more likely, we now have a necessary condition for states to believe that mutual cooperation has higher payoffs than mutual defection [@keohane_reciprocityinternationalrelations_1986].

Lastly, even in asymmetrical alignments where the strong state is determining the terms of the agreement, both states are able to leverage the power of their allies to achieve international outcomes that are in their favor [@davidson_americaallieswar_2011]. Smaller states may desire institutionalizing their relationship with more dominant states precisely because that increases their bargaining leverage and creates mutual interdependence [@bosse_bargainingpoweralliance_2010; @schneider_weakstatesinstitutionalized_2011]. This provides a way for both actors in an alignment relationship to value the alliance independent of the degree of control they exercise in determining the structure and terms of that alignment [@schroeder_alliances18151945_2004; @weitsman_dangerousalliancesproponents_2004; @bearce_alliancesinternalinformation_2006].

Hierarchy also reduces the cost of coordination because it improves information processing [@chandler_visiblehandmanagerial_1977; @gulati_twofacetscollaboration_2012], ensures actors know what communication is authoritative [@galbraith_organizationdesign_1977], and simplifies decision-making [@chandler_visiblehandmanagerial_1977]. "The focus shifts to creating structures, institutions, and relationships that enable partners to work together across boundaries. The coordination perspective emphasizes organization design, communication, and process management as requisite skills of alliance managers"[@gulati_twofacetscollaboration_2012, p 533]. This helps produce things like standard operating procedures (SOP), unified command structures, and authoritative rules and procedures that create the type of task coordination that is needed for certain military strategies and structures [@march_organizations_1958]. These help minimize communication, simplify decision-making, reduce uncertainty about future tasks, and prevent disputes [@pondy_otherhandclapping_1977]. By reducing the costs of coordination, institutionalization makes the interdependence of tasks easier which, in turn, facilitates a division of labor [@axelrod_achievingcooperationanarchy_1985]. Institutionalization allows actors to figure out the "anticipated organizational complexity of decomposing tasks among partners along with ongoing coordination of activities to be completed jointly or individually across organizational boundaries and the related extent of communication and decisions that would be necessary" [@gulati_architecturecooperationmanaging_1998, 304].

*Hypothesis: States in hierarchical cooperative security alignments should have a higher division of labor than states in non-hierarchical cooperative security alignments.*

Hierarchy is a way of institutionalizing cooperation because decision-making becomes relatively centralized based on the preferences of asingle actor. Policymakers have recognized the importance of institutionalization of security relationships, with US Navy Rear Admiral @smith_strategiccooperationeverybody_2013 noting "what is unavoidably true is that, in the absence of an institutionalized habit of pooling our naval resources in steady-state planning, the best of intentions will not result in meaningful implementation of a cooperative strategy." The high degree of institutionalization in bodies like NATO have "reduced uncertainties about each other's behavior and provided mechanisms for the peaceful resolution of disputes" [@cottey_eastcentraleurope_1995, 6]. So the issue is not just whether or not states have a formal alliance. NATO, SEATO, CENTO, and ANZUS all involve actors with similarly aligned interests but these institutions have varying degrees of institutionalization in their decision-making structures. More institutionalized alliances like NATO should have higher degrees of specialization and complementarity than less institutionalized ones like SEATO.

I predict no division of labor by either state in cooperative security relationships that are anarchic (the absence of a cooperative security relationship). The outcome of bargaining over collective defense is too costly and thus unresolved. In the absence of strategic compatibility or hierarchy in that relationship, the expected costs of opportunism and coordination exceed the expected gains of cooperation [@wendt_hierarchyanarchyinformal_1995]. In this situation, functionally differentiating one's military by specializing in a way that creates a dependence on other states leaves one vulnerable to costly and likely opportunistic behavior by others [@lake_risefallfuture_1997]. As a result, there will be no functional differentiation outside that explained by differences in factor endowment and geography. Instead, states occupying similar positions will adopt similar strategies for power and security.

# Empirics {#sec-empirics}

## Dependent Variable {#sec-dependent-variable}

The dependent variable is the division of defense labor among allied states.[^12] I conceptualize security as an output that requires a number of distinct tasks (observed as military capabilities) that can, in theory, be distributed among a number of members of an alliance [@gorelick_normalizedmutualentropy_2004]. This can be observed as a matrix where each row is an alliance member and each column is a functional security capability. Each cell represents the quantity of that technology owned by that alliance member. A division of labor can then be quantified as the degree to which each member of the alliance specializes in one activity or performs all tasks, whether a task is performed by one alliance member or many of them, and what activities are performed together by the same individual [@gorelick_quantifyingdivisionlabor_2007]. I assume that these technologies could at least in theory be allocated to the defense of other allied states.[^13] When multiple allies possess the same military capabilities and omit the same military capabilities, their division of labor is low and can be described as redundant -- neither is making a substantial unique contribution to their "pooled" defense capabilities. By comparison, when multiple allies possess different military capabilities from one another, they each fill in the gaps such that the combination of their capabilities is distinct from, and more well-rounded, than each individual state.

[^12]: @gannon_alliesarmamentsexplaining_2023 has examined the role of alliances in explaining military specialization. While specialization is a characteristic of an individual state, division of labor is a characteristic of a group. Group members can all be specialized without that necessitating a division of labor if they are all specialized in the same tasks.

[^13]: A future extension would account for how factors like geography and basing mean some capabilities are more readily available to allies than others. For now, the rate of availability is assumed to be greater than 0 and constant across all technologies.

The division of labor between states in an alliance is calculated as their 'niche width' which measures the weighted pairwise similarity of their military portfolios in a given year [@bolnick_measuringindividuallevelresource_2002]. For each year $t$, considering an $n \times m$ matrix for every alliance member $N$ and technology $M$. The pairwise similarity measure $\theta_{ij}$ = $\sum_{m}^{}min(p_{im}p_{jm})$ for every states $i$ and $j$ where $p_{im}$ and $p_{jm}$ represent their respective proportions of technology $m$ [@zaccarelli_rinsppackageanalysis_2013]. The measure is normalized between 0 and 1 where 0 means the group of states have entire dissimilar militaries and 1 means they have the exact same technologies in identical proportions. A division of labor can thus be observed as the complementarity associated with high dissimilarity since it means your partners possesses capabilities you do not and visa versa. @fig-dv illustrates the distribution of the division of labor scores in the data.

```{r}
#| label: fig-dv
#| fig-cap: "Distribution of alliance-year division of labor. The measure is bounded between 0 and 1, with 1 representing the highest theoretically possible level of division of labor."
#| fig-pos: 'h'

df %>%
  ggplot(aes(x = dol_nicheolap)) +
  geom_histogram(bins = 30, color = "black", fill = "gray") +
  geom_vline(aes(xintercept = mean(dol_nicheolap, na.rm = TRUE)), 
             linetype = "dashed", linewidth = 0.6) +
  annotate("text",
           x = mean(df$dol_nicheolap, na.rm = TRUE) + 0.1,
           y = 200,
           label = paste("Mean =", round(mean(df$dol_nicheolap, na.rm = TRUE), 3))) +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  labs(title = "Division of Labor Index across Alliance-Years",
       x = "Division of Labor",
       y = "Number of alliance-years") +
  theme_classic() +
  theme(text = element_text(size = 10))
```

The measure is weighted, meaning that their similarity is considered proportional to the size of each state (measured as the number of technologies it possesses) as well as the abundance of each technology. For example, two states both possessing 100 main battle tanks does not contribute very much to their similarity because main battle tanks are quite common. By contrast, the possess of ICBMs by two states would contribute much more to their similarity given the rarity of that capability. This measure is appropriate since it can account for actors that possess 0 of a given technology and was developed to account for wide differences in the availability of each technology.[^14]

[^14]: This measure and similar measures were initially developed as an ecological index to compare biodiversity across different sites. See @bolnick_measuringindividuallevelresource_2002, @jeanson_emergencedivisionlabour_2005, @jeanson_emergenceincreaseddivision_2007, @dornhaus_specializationdoesnot_2008, @dormann_indicesgraphsnull_2009, @zaccarelli_rinsppackageanalysis_2013, @leroy_complementarityrarityspecialisation_2014, and @holbrook_individualdifferencespersonality_2014.

Cooperation between the US and European countries during the Iran-Iraq Tanker War worked because each was *uniquely* specialized in a way that produced complementarity rather than redundancy. The US provided large surface vessels that the other states did not have and the Netherlands and Belgium provided minesweepers that the US did not have. If the US and European countries had all been specialized in minesweepers, this cooperative security arrangement would not have made sense. As another example, while the US possesses only 2 Arctic-capable icebreakers (as opposed to Russia's 40), 7 of the 8 Arctic nations are US allies via NATO or NATO-partners [@markowitz_perilsplentyarctic_2020, p 76-78]. For example, Thule Air Base houses the US Ballistic Missile Early Warning System (BMEWS), yet winter access to the base by sea is provided entirely by Canada's icebreaking fleet [@cross_natoalliancerole_2019]. By having a division of labor whereby US allies operate icebreakers in the Arctic, "allies and partners can free up U.S. time and resources to focus elsewhere. They can also help improve situational awareness and manage tensions more broadly to minimize dangers and create opportunities in and near the North American and European Arctic" [@avey_icebreakergapdoesn_2019]. These capabilities thus complement the technological omission of the United States.

## Independent Variable and Controls {#sec-independent-variable-and-controls}

The explanatory variables concern variation in alliance relationships. I differentiate alliances using the two variables described earlier -- strategic compatibility and hierarchy. The combination of the two variables provides both across- and within- alliance comparisons. Alliances with high strategic compatibility and high hierarchy should have the highest division of labor. I define these variables at the alliance level for the network of states that share a defensive alliance pact. Data on alliance pact membership comes from the Alliance Treaty and Provisions (ATOP) data set version 5 [@leeds_alliancetreatyobligations_2002].

Strategic compatibility describes the consistency of two states' security interests and agreement on the nature of the international threat environment. I create a new index of the strategic compatibility across alliance members is a given year based off of a similar measure used by @leeds_terminatingallianceswhy_2007 and @poast_arguingalliancesart_2019.[^15] I first create a measure of each country-year's threat environment, where threats are defined by a combination of power and foreign policy orientation. State B is considered a part of State A's threat environment if it meets any of the following conditions: (1) State B is a non-allied great power, (2) State B is a strategic rival, as defined by @thompson_identifyingrivalsrivalries_2001 and @colaresi_strategicrivalriesworld_2008, or (3) State B is contiguous with State A and has an s-score that is below the population median [@leeds_terminatingallianceswhy_2007]. From this, I then identify the shared threat environment across alliance members in a given year, where the shared threat environment is defined as the ratio of summed CINC scores for common threats divided by all threats squared [@poast_arguingalliancesart_2019, 55]. The resulting index is thus highest for alliances where all members have the same threat environment and lowest for alliances where the common threat(s) the members face are much less militarily salient than the threats they each face individually.

[^15]: While other work has used conflict history as a proxy for interest alignment, it is an inappropriate measure for strategic compatibility given its endogenous relationship with alliance military strength [@leeds_alliancesdeteraggression_2003].

Operationalizing hierarchy in international politics has been a notoriously challenging enterprise [@macdonald_rolehierarchyinternational_2008]. Recent methodological advancements have benefited from the turn towards network analysis in international relations which provides a better picture of true interstate relations than previous dyadic measures [@poast_misusingdyadic_2010; @dorff_networksdyadssocial_2013; @cranmer_critiquedyadicdesign_2016; @poast_dyadsaredead_2016]. Building off @beardsley_hierarchyprovisionorder_2020's innovative measure of hierarchy as the network centrality of joint-production security communities, I consider the degree of influence that each state in an alliance network has on its security community, where influence is observed as relative dependency in directed arms sales. Using SIPRI arms sales data, I create a global network with weighted directed edges accounting for arms sales between each pair of states. Subsetting that to each alliance network, I then measure hierarchy as the difference in importance of each state within an alliance network. In an alliance network where all states are equally important, the degree of hierarchy can be described as low. But if one state is highly influential in an alliance network, the degree of hierarchy is high because there is a relational power imbalance as a structural property of the network of states [@kahler_networkedpoliticsagency_2009]. The importance of each state is measured as the drop in the network's Laplacian energy when that vertex is removed, where Laplacian energy is quanitified as the sum of squares of the eigenvalues in the Laplacian matrix of arms sales [@qi_laplaciancentralitynew_2012; @boley_randomwalklaplacian_2018; @cordeiro_dynamiclaplaceefficient_2018]. This produces a Laplacian measure of importance/centrality for each state within an alliance. For each alliance year, I then take the mean between the highest Laplacian value and every other Laplacian value which creates a network-level measure of hierarchy. This measure has been previously used to identify central actors in terrorist networks, which validates it as an accurate measure of the relative importance of political actors [@qi_terroristnetworksnetwork_2013].

The models include control variables for geography, economic capacity, and the threat environment since existing theories suggest that these factors could influence the degree to which two states have a similar distribution of military capabilities or that could influence the degree to which those states cooperate on security issues.[^16] More militarily powerful alliances may have a different division of labor, the model includes a control for the squared total CINC score among all alliance members [@fordham_allalliancesare_2016]. The model also controls for the proportion of states in an alliance that are a democracy, measured as a Polity score greater or equal to 6, since democratic allies may trust each other to different degrees [@chiba_carefulcommitmentsdemocratic_2015]. Since geographic proximity may better enable allied states to cooperate militarily, the model controls for the logged maximum distance between alliance members [@bak_allianceproximityeffectiveness_2018] as well as the proportion of alliances members that are contiguous [@fordham_allalliancesare_2016]. Larger alliances may inherently have a lower division of labor since more partners increases the potential for overlap and redundancy in capabilities, so the model includes a control for the logged number of alliance members in a given year [@chiba_carefulcommitmentsdemocratic_2015]. Established alliances are more likely to have deeper roots for cooperation and thus have an easier time cooperating, so the model controls for the average number of years each member has been in the alliance [@benson_assessingvariationformal_2016]. The model also controls for time using cubic year polynomials, as changes in the quantities and qualities of technologies over time may impact the observed division of labor [@carter_backfuturemodeling_2010].

[^16]: Many of these variables were created using the peacesciencer R package [@miller_peacesciencerpackagequantitative_2022].

## Model and Results {#sec-model-and-results}

The dependent variable is the division of labor of military capabilities measured for each alliance-year. The dependent variable is continuous and bounded between 0-1 where 0 represents redundant militaries and 1 represents complementary militaries that constitute a division of labor. The models are estimated using a series of ordinary least squares (OLS) regressions with standard errors clustered at the alliance level.

Table 1 shows the results of a series of models, with varying specifications. I first estimate a model using just the two explanatory variables of interest. Without control variables, strategic compatibility and hierarchy both have a statistically significant positive association. Model 3 considers both independent variables and finds similar results. The fully specified model (4) includes all control variables and produces similar results suggesting that strategic compatibility and hierarchy are positively associated with a higher division of labor. I do not interpret the model coefficients for any of the control variables, due to the absence of confounders for those variables [@dworschak_biasmitigationempirical_2023].

```{r model}
df_m1 <- df %>%
  dplyr::mutate(members_log = log(members),
                strcomp_cinc_totalsq = strcomp_cinc_total^2,
                hier_laplacesq = hier_laplace^2) %>%
  dplyr::select(dv = dol_nicheolap,
                strcomp_cinc_totalsq,
                hier_laplacesq,
                cinc_totalsq,
                demo_prop,
                dist_maxlog,
                contig_prop,
                members_log,
                allianceage_avg,
                year_norm,
                year_sq,
                year_cube)

m1_bi <- lm(dv ~ strcomp_cinc_totalsq, data = df_m1)
m1_bicse <- lmtest::coeftest(m1_bi, vcov = sandwich::vcovCL(m1_bi, factor(df$atopid)))

m2_bi <- lm(dv ~ hier_laplacesq, data = df_m1)
m2_bicse <- lmtest::coeftest(m2_bi, vcov = sandwich::vcovCL(m2_bi, factor(df$atopid)))

m3_bi <- lm(dv ~ strcomp_cinc_totalsq + hier_laplacesq, data = df_m1)
m3_bicse <- lmtest::coeftest(m3_bi, vcov = sandwich::vcovCL(m3_bi, factor(df$atopid)))

m4 <- lm(dv ~ ., data = df_m1)
m4_cse <- lmtest::coeftest(m4, vcov = sandwich::vcovCL(m4, factor(df$atopid)))

models <- list(m1_bicse, m2_bicse, m3_bicse, m4_cse)

coefs <- c("strcomp_cinc_totalsq" = "Strategic Compatibility",
           "hier_laplacesq" = "Hierarchy$^2$",
           "cinc_totalsq" = "CINC$^2$",
           "demo_prop" = "Democracy Ratio",
           "dist_maxlog" = "Max distance (log)",
           "contig_prop" = "Contiguity Ratio",
           "members_log" = "Alliance Members (log)",
           "allianceage_avg" = "Alliance Age (avg)",
           "year_base" = "Year",
           "year_sq" = "Year$^2$",
           "year_cube" = "Year$^3$")

modelsummary::modelsummary(models,
                           statistic = 'p.value',
                           stars = TRUE,
                           coef_map = coefs,
                           escape = FALSE,
                           output = "markdown") %>%
  kableExtra::add_footnote(label = "All models include country-clustered standard errors.")
```

These results are robust to a series of alternate model specifications provided in the appendix. Further models operationalize strategic compatibility using other commonly used metrics like s-scores [@signorino_taubnottaub_1999; @leeds_allianceinstitutionalizationalliance_2005] and joint alliances, as well as operationalizing hierarchy as signaled support by major powers [@mcmanus_identifyinglevelmajor_2019], and include alternate measures of geography like distance [@weidmann_geographyinternationalsystem_2010]. As the dependent variable is bounded between zero and one, a more appropriate model specification than OLS is a zero-inflated beta model, which is also included in the appendix and produces results that are substantively similar.

A more intuitive interpretation of the two independent variables is shown in Figure TBD. The first independent variable -- strategic compatibility -- is plotted on the X-axis. The partial residual are shown based on different values of hierarchy (low and high on its original ordinal scale). Division of labor is highest in alliances that have high strategic compatibility and high hierarchy and lowest in alliances that have low strategic compatibility and low hierarchy. Alliances high in one of the two independent variables but not the other have a division of labor somewhere in between, with no clear indication about which of these matters more. The differences in slope are the result of a multiplicative interaction term between the two independent variables, although it is statistically insignificant.

**Add Figure 3**

This relationship is also substantively significant. A one standard deviation increase in strategic compatibility is associated with an increase in division of labor of just over a tenth of a standard deviation. Put concretely, this is the difference between the military division of labor between the United States and Japan in 1984 and in 2000. For hierarchy, a shift from low hierarchy to high hierarchy is associated with an increase in division of labor of roughly one quarter of a standard deviation. This can be represented by the difference in the division of labor between the United States and Poland and the United States and Ukraine in 2014. Poland, a NATO member, has a higher division of labor with the US than does Ukraine, a non-NATO member. Poland has accomplished this by specializing in short-range transport aircraft and surface to air (SAM) missile defense systems that the United States does not possess in particularly high quantities. This is not coincidental, as Poland learned after the Bosnian war that their contribution to NATO air forces was marginal at best and they were instead better off contributing ground troops and short-range transport aircraft [@szayna_natoenlargement20002015_2001; @wolosz_wingswhiteeagle_2004]. Their first formal integration with NATO defense forces involved air defense systems as NATO "encouraged Poland to acquire more of a niche capability within the alliance, which would antagonize Russia less, but would also fit better into collective defense efforts (a smaller amount of mobile and highly technological units that could be deployed abroad quickly were encouraged.)" [@burton_natodurabilitypostcold_2018, p 38].

## Discussion {#sec-discussion}

Pessimism about states' ability to cooperate on security issues under anarchy is challenged by the reliance seen under some alliances. Contrary to claims that self-help prevents states from strategically choosing to functionally differentiating their defense capabilities, various forms of governance in interstate relations can allow functional differentiation in defense based on the presence of strategic compatibility and hierarchy.

NATO expansion provides an illustrative example.[^17] In 1994, Albania and Georgia were among the 18 Former Soviet Union (FSU) and Warsaw Pact states to join the Partnership forXXPeace, a new NATO program designed to initiate military cooperation with prospective applicants [@szayna_natoenlargement20002015_2001]. Albania and Georgia were similar in many respects like GDP, military spending, geographic size, and coastline.

[^17]: For a more detailed account of the NATO expansion and Former Soviet Union and Warsaw Pact cases, see [@gannon_securitycooperationcase_2021].

Figure 3 illustrates how this case maps onto the results of the statistical model. In 1994, both Albania and Georgia had similar divisions of labor with the United States (0.6). Albania's division of labor with the United States increased after they joined the Membership Action Plan (MAP) and again after they were granted NATO membership. By 2014, their division of labor with the United States was around 0.75. Georgia, by comparison, experienced a relatively constant decrease in their division of labor with the United States, with that decrease sharpening most following the 2008 war with Russia and consequent lack of confidence in their participation in the NATO alliance.

The difference in their respective divisions of labor with the United States can be attributed to Albania's decision to specialize their military while Georgia opted for a more conventional (small) full-spectrum force. Figure XX shows Albania and Georgia's military portfolios as a percent of the world's share for each capability. While not identical, the two states were fairly similar in having moderate navies, smaller but capable land forces, and very limited air capabilities. Figure XX shows how that changed a decade later. Albania specialized in naval patrol vessels with a range that allowed them to patrol as far as Portugal, despite their own coastline being only 225 miles [@polak_natomembershipalbania_2009]. They also omitted air capabilities almost entirely, disposing of all former Soviet combat helicopters and fixed wing aircraft in favor of minimal air power capable only of surveillance and humanitarian missions. Choosing to these omissions and overproductions were consistent with their doctrinal aim of performing the niche role of maritime security for its NATO allies. Albania even went so far as to name their 2004 Military Strategy document "Strategy of NATO-integration" which called for a shift to sea and air surveillance and humanitarian assistance so that it could fulfill its goal of "providing combat readiness for defence; monitoring and surveillance of Albania's sea, air, and land territorial space; participating in humanitarian assistance and disaster relief operations; fighting terrorism; and participating in peace support missions" [@governmentoftherepublicofalbania_militarystrategyrepublic_2004].

Georgia, on the other hand, had received only weak promises of support from NATO in the lead up to the 2008 war with Russia [@lanoszka_tangledrosetheories_2017]. After Russia's attack -- purportedly motivated by Russian concerns about NATO expansion -- NATO confirmed that Georgia would not invited to NATO at this time, although it would be possible at some point in the future. While the 2011 National Security Concept of Georgia mirrors the 2005 version in discussing the importance of NATO membership for Georgia's defense, there is pessimism about NATO's willingness to defend Georgia [@merabishvili_perceptionnationalsecurity_2016]. Georgian President Saakashvili remarks that "NATO will not help us in this" [@russianagenciessaakashvili_2008] occurred a month after US ambassador John Tefft told Saakashvili that "Tbilisi, if it acted, would stand alone" [@asmus_littlewarthat_2010, p 144]. As a result, Georgia has not been able to mirror Albania in downsizing their air force and instead specializing in niche maritime patrol capabilities that contribute to the common defense. Instead, they have continued to diversify their military portfolio with an eye toward self-defense against Russian territorial aggression, as evidenced by their more capable land and anti-air capabilities. Current debates have shifted from whether NATO expansion deterred Russian aggression [@lanoszka_thankgoodnessnato_2020; @shifrinson_eastboundunitedstates_2020] to how NATO membership shapes the form of Russian aggression [@gannon_shadowdeterrencewhy_2023]. These findings help us further advance our understanding of these issues by identifying how NATO membership shapes the composition of military capabilities that states possess.

# Conclusion {#sec-conclusion}

The primary purpose of a state's military is to improve their security. Despite a recognition that this is conditioned by considerations like economic and geographic constraints and differences in what security threats state face, military power is still treated as a fungible asset that varies in size, but not in composition. However, states differ in what military capabilities they choose for their security. When states with similar economic and geographic constraints choose different force structures it may be because the optimal force structure for is conditioned by a state's cooperative security relationships. States do not just decide between internal and external balancing. External balancing influences how a state internally balances because what military capabilities a state needs is a function of the military capabilities their ally possesses.

These findings also point to a new mechanism by which states can prevent opportunism by alliance partners. Conventional wisdom holds that asymmetric alliances have trouble with reliability-enhancing features like precision, issue linkage, and institutionalization since the larger state does not need reliability enhancement and the smaller state cannot get it [@mattes_reputationsymmetryalliance_2012]. Dominant states want to free ride on their smaller partners, but cannot because they have more at stake and thus end up over-providing [@olson_economictheoryalliances_1966]. Rather than coercing their allies into contributing, allies can engage in a strategic division of labor where each provides useful capabilities in a way that is incentive compatible for both partners. Specialization is thus a way of preventing opportunism by limiting adventurism by smaller states (resolving entrapment) and preventing abandonment by the larger state.

Convergence of foreign policy preferences and institutionalized hierarchy interact to shape the type of military capability portfolio a state maintains. Among other purposes, interstate alignments help a state defend itself better than they could defend themselves alone. However, there are downsides to relying on other states for defense. Other states could behave opportunistically by defecting in a way that presents a risk to your national security and defense and second, coordinating that cooperation can be costly. As a result, alignments must contain ways to guard against the risk of opportunism and costs of coordination [@yarbrough_cooperationgovernanceinternational_2016]. This insight can help inform current debates about changing NATO relations and identify the consequences of allies trusting each other less than they used to. These debates often turn to the question about whether allies are contributing enough to the alliance. But by looking at what states are contributing to the common defense, rather than how much they are spending, new perspectives on burden-sharing and the value of the alliance may emerge. After all, the composition of military assets, not just the amount spent, is what is truly of tremendous consequence for how NATO deals with future threats.

By applying economic and business organization theories about patterns of production across actors in the same space, a theory of a shared production model of military capabilities identifies a way that states can get the benefits of specialized production -- economic, political, and military -- while minimizing the costs of omitting some assets while overproducing others. In this way, states strategically choose to functionally differentiate through interstate security cooperation when it is conducive to a division of labor across nations. This happens when intra-alliance bargaining can overcome the costs of opportunism and coordination that otherwise inhibit reliance on others in the high stakes realm of the security and survival of the state. These two problems can be overcome when the states in question have (1) closely aligned interests and (2) there is an institutionally-designed command and control system. When states are able to cooperate over security issues, that cooperation manifests itself in specialized and complementary military capabilities.

\newpage

# References {#sec-references}
